PCA

### 2023-11-28 DRAFT_PCA.Rmd
Formatting updates to PCA draft, including equation setting

```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```
Authors: KE Lotterhos, DL Davenport, S Truskey <br>
Acknowledgements/ Page Reviewers: Jason Johns

@@

* [What is PCA and what are its common uses in population genetics?](#whatspca)
* Considerations before sampling:
    + [How one chooses to sample from a landscape](#sampling)
    + [Sample sizes](#samplesize)
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + [Batch effects](#batcheffects)
    + [LD and inversion effects](#ld)
    + [Missing data and imputation](#missingdata)
    + [Hybridization and inbreeding effects](#hybrid)
    + [Horseshoe and wave patterns](#horseshoe)
* Comparison to other common approaches to investigating population genetic structure
    + [PCA  vs ADMIXTURE/STRUCTURE](#pca_admix)

@@

Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) . It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see @Pearson1901-xe). Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see @Thioulouse2018-bj).  In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.

The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs). When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e. often more variables than observed entities, n << m, see this [YouTube playlist](https://www.youtube.com/playlist?list=PLMrJAkhIeNNSVjnsviglFoY2nXildDCcv) for details). Other methods use the covariance (or correlation) matrix. The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).

The eigenvalues represent the amount of variance explained per component and are typically represented as a barplot, in descending order (Figure 1) (also called variable loadings). The percentage of explained variance by each eigenvector will vary between datasets since it is highly dependent on the size of the data and the correlation structure between variables. The eigenvalues are typically used as an empirical way of choosing the number of PCs to represent the data. Usually, the “elbow” of this barplot shows the number of the least amount of variables (PCs) explaining the most amount of variation.

@@

To get a better understanding of PCA, we recommend reading these useful reviews and studies of PCA:

@@

**Number of SNPs**
The number of SNPs sampled has a large impact on the resolution of populations in PC space. In PCA, there is a critical signal-to-noise ratio below which the true structure of the signal cannot be recovered - in other words for genotype data, as genetic distance (i.e. FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases. The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms):

FST > 1/sqrt(number_of_SNPs * number_of_samples_per_group) 		Eq1
Rearranging, if we know FST and the number of samples per group, we can calculate how many SNPs we would need to sequence for our PCA to be above the signal-to-noise ratio; or if we know FST and the number of SNPs, we can calculate how many samples per group we would need to sequence for our PCA to be above the signal-to-noise ratio:

Number_of_samples_per_group <- 1/(number_of_SNPs * FST^2)		Eq2

Number_of_SNPs <- 1/(Number_of_samples_per_group * FST^2)		Eq3

For instance, if FST = 0.01 and there is 100 samples per group, then you would need 1/(100 * 0.01^2) = 100 SNPs . If FST = 0.001 and there is 20 samples per group (not out of the question for a marine species), then you would need 1/(20 * 0.001^2) = 50,000 SNPs. Anecdotally, we have observed with SNP data for a high gene flow marine species (low FST) that >50,000 SNPs was needed to resolve structure in PC space between adjacent populations.

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 4. Figure 2 from MacVean - Simulated a population (Fig 2A), where the expected locations for the two populations on the first PC are defined by the time-since-divergence (the lines in Fig 2C). The signal-to-noise threshold for a sample size of 100 and FST=0.01 is 100 SNPs (see Eq3 above). Fig 2C shows that around or below this threshold of 100 SNPs, the PCA is affected by random clustering. The author notes that the separation of samples (Fig 2C) with 10 SNPs does not correlate with population divergence, but reflects random clustering arising from the small numbers of SNPs."}

@@

MacVean (2009) used simulations of two populations of equal size to show how the number of individuals sampled per group influenced PC projections (Figure 2a, MacVean, 2009). Figure 2B (MacVean, 2009) shows how groups with larger sample sizes will lie closer to the origin of the first PC axis, while Figure 2c shows the effect of using fewer SNPs on the inferring the true population configuration, where lines indicate the expectation of population scores on the PC-axis.

The author extended this analysis to a 9-population stepping stone lattice. Figure 3 from the paper shows how differences in the number of samples per population can warp the projection space of PCA, even when migration rates and effective population size in each deme are the same. Note that FST among demes would be similar in all panels because it is based on an allele frequency in each deme (although small differences would occur due to sampling error when fewer individuals within a deme are sampled).

@@

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing genotypes become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). @Yi2022-bi demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

@@

Imputation via the mean genotype within a site, sampling location, or population. 
In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population. The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 12). This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values. This method is not recommended (see @Yi2022-bi), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:----------|:------------------------------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |@wold1975path|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
 |@purcell2007plink|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|

>>>>

```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```
Authors: KE Lotterhos, DL Davenport, S Truskey <br>
Acknowledgements/ Page Reviewers: Jason Johns, Katherine Silliman

@@

* [What is PCA and what are its common uses in population genetics?](#whatspca)
    + [Recommended reading](#recreading)
* Considerations before sampling:
    + [How one chooses to sample from a landscape](#sampling)
    + [Sample sizes](#samplesize)
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + [Batch effects](#batcheffects)
    + [LD and inversion effects](#ld)
    + [Missing data and imputation](#missingdata)
    + [Hybridization and inbreeding effects](#hybrid)
    + [Horseshoe and wave patterns](#horseshoe)
* Comparison to other common approaches to investigating population genetic structure
    + [PCA  vs ADMIXTURE/STRUCTURE](#pca_admix)

@@

Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka **variance**) . It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see @Pearson1901-xe). Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see @Thioulouse2018-bj).  In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.

The output of PCA are the eigenvalues and eigenvectors (**eigenvalues** = principal components (PCs), **eigenvectors** = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs). When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called **singular value decomposition (SVD)** or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e. often more variables than observed entities, n << m, see this [YouTube playlist](https://www.youtube.com/playlist?list=PLMrJAkhIeNNSVjnsviglFoY2nXildDCcv) for details). Other methods use the covariance (or correlation) matrix. The data are usually **centered** (so it becomes zero mean gaussian, center = TRUE ), and **scaled** ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).

The eigenvalues represent the amount of variance explained per component and are typically represented as a barplot, in descending order (Figure 1) (also called **variable loadings**). The percentage of explained variance by each eigenvector will vary between datasets since it is highly dependent on the size of the data and the correlation structure between variables. The eigenvalues are typically used as an empirical way of choosing the number of PCs to represent the data. Usually, the “elbow” of this barplot shows the number of the least amount of variables (PCs) explaining the most amount of variation.

@@

### Recommended reading{#recreading} 
To get a better understanding of PCA, we recommend reading these useful reviews and studies of PCA:

@@

**Number of SNPs**
The number of SNPs sampled has a large impact on the resolution of populations in PC space. In PCA, there is a critical signal-to-noise ratio below which the true structure of the signal cannot be recovered - in other words for genotype data, as genetic distance (i.e. FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases. The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms). Note that number of samples below represents haploid samples, so double the number for diploid samples:

**Eq. 1**  
$$FST > \frac{1}{\sqrt{(number\_of\_SNPs * number\_of\_samples\_per\_group)}}$$ 	
Rearranging, if we know FST and the number of samples per group, we can calculate how many SNPs we would need to sequence for our PCA to be above the signal-to-noise ratio; or if we know FST and the number of SNPs, we can calculate how many samples per group we would need to sequence for our PCA to be above the signal-to-noise ratio:

**Eq. 2**
$$number\_of\_samples\_per\_group > \frac{1}{(number\_of\_SNPs * FST^2)}$$
**Eq. 3**
$$number\_of\_snps> \frac{1}{(number\_of\_samples\_per\_group * FST^2)}$$

For instance, if FST = 0.01 and there is 100 samples per group, then you would need at least 1/(100 * 0.01^2) = 100 SNPs . If FST = 0.001 and there is 20 samples per group (not out of the question for a marine species), then you would need 1/(20 * 0.001^2) = 50,000 SNPs. Anecdotally, we have observed with SNP data for a high gene flow marine species (low FST) that >50,000 SNPs was needed to resolve structure in PC space between adjacent populations.

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 4. Figure 2 from MacVean (2009) - Simulated a population (Fig 2A), where the expected locations for the two populations on the first PC are defined by the time-since-divergence (the lines in Fig 2C). The signal-to-noise threshold for a sample size of 100 and FST=0.01 is 100 SNPs (see Eq3 above). Fig 2C shows that around or below this threshold of 100 SNPs, the PCA is affected by random clustering. The author notes that the separation of samples (Fig 2C) with 10 SNPs does not correlate with population divergence, but reflects random clustering arising from the small numbers of SNPs."}

@@

MacVean (2009) used simulations of two populations of equal size to show how the number of individuals sampled per group influenced PC projections (Figure 4A). Figure 4B shows how groups with larger sample sizes will lie closer to the origin of the first PC axis, while Figure 4C shows the effect of using fewer SNPs on the inferring the true population configuration, where lines indicate the expectation of population scores on the PC-axis.

The author extended this analysis to a 9-population stepping stone lattice. Figure 3 from the paper shows how differences in the number of samples per population can warp the projection space of PCA, even when migration rates and effective population size in each deme are the same (Figure 5). Note that FST among demes would be similar in all panels because it is based on an allele frequency in each deme (although small differences would occur due to sampling error when fewer individuals within a deme are sampled).

@@

**Imputation via the mean genotype in the metapopulation.**

In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing genotypes become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). @Yi2022-bi demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 9).

@@

**Imputation via the mean genotype within a site, sampling location, or population.**

In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population. The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 10). This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values. This method is not recommended (see @Yi2022-bi), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.

@@

| Method or Function, command call | Treatment of Missingness | Ref |
|:----------|:------------------------------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |@wold1975path|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|@purcell2007plink|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|
